Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-06-22 paper/paper1/ms.tex
re-enable format-on-save with latexindent, flaky on linux tho...
One less placeholder
Minor paper edit and nice formatting

\documentclass[modern]{aastex631}
\usepackage{hyperref}
\bibliographystyle{aasjournal}
%\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}

@@

In extreme precision radial velocity (EPRV) applications, cross-correlation methods work \citep{2018A_A...620A..47D}, but have many limitations \citep{2022arXiv220110639Z}.  Addressing telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption before convolution with an instrumental kernel.  This telluric joint modeling capability does not yet exist at a precision that can meet these strenuous demands.

@@

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \cite{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  We show how to combine both stellar and telluric models in a joint modelling approach.  In Section \ref{transferLearn} we describe how to fine-adjust the cloned models to real-world data from high-grasp near-IR spectrographs, using a transfer-learning step.
%Section \ref{secInjRec} exhibits injection-recovery tests on simulated data of known properties, revealing the scalability and fundamental limits of the technique.  
Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \ref{secDiscuss}.

@@

At this stage, we have the choice of whether to work in linear or log scale flux units. Adopting the log of the flux would ensure that the cloned model possesses only positive flux values, a desirable trait of any physical spectral model. We have implemented both modes in \texttt{blas\'e}, allowing users to choose their preference.  We only narrate the linear flux unit description in the main text of this document, for the sake of clarity, and since most practitioners tend to think of flux in terms of linear flux units.  The data-model comparison step will always take place in linear flux units, so the only operational difference is the behavior for deep and saturated lines.  Appendix \censorbox{XX} lists the equations adjusted to log flux units.

@@

% Table
\begin{deluxetable}{cp{10cm}}
    \tabletypesize{\scriptsize}
    \tablecaption{Notation used in this paper\label{table2}}
    \tablehead{
        \colhead{Symbol} & \colhead{Meaning}
    }
    \startdata
    \hline
    \multicolumn{2}{c}{Spectra}\\
    \hline
    $\bm{\lambda}_S$ & Native wavelength coordinates of the precomputed stellar spectrum\\
    $\bm{\lambda}_T$ & Native wavelength coordinates of the telluric spectrum\\
    $\bm{\lambda}_D$ & Native wavelength coordinates of the data spectrum\\
    $\mathsf{S}_{\rm abs}$ & Flux values of the precomputed synthetic stellar spectral model $\bm{\lambda}_S$\\
    $\mathsf{B}$ & Blackbody of temperature $T_{\mathrm{eff}}$ to coarsely normalize $\mathsf{S}_{\rm native}$\\
    $\mathsf{P}$ & Smooth polynomial to refine continuum-normalization\\
    $\mathsf{S}$ & Continuum normalized augmentation of $\mathsf{S}_{\rm abs}$\\
    $\mathsf{T}$ & Transmission values of the precomputed synthetic telluric model \\
    $\mathsf{D}$ & The observed data spectrum flux values\\
    $\bm{\epsilon}$ & The estimated uncertainties in the data spectrum\\
    $\mathsf{S}_{\rm clone}$ & Evaluable and tunable cloned flux model of $\mathsf{S}$\\
    $\mathsf{T}_{\rm clone}$ & Evaluable and tunable cloned transmission model of $\mathsf{T}$\\
    $\mathsf{S}_{\rm ext}$ & An augmentation of $\mathsf{S}_{\rm clone}$ with $v\sin{i}$ convolution and $RV$ translation\\
    $\mathsf{M}_{\rm joint}$ & The joint stellar and telluric model: $\mathsf{S}_{\rm ext} \odot \mathsf{T}_{\rm clone}(\bm{\lambda}_S)$  \\
    $\mathsf{M}$ & Joint model convolved
    with instrumental kernel and resampled to $\bm{\lambda}_D$\\
    $\mathsf{R}$ & The residual spectrum between a pair of inputs, \emph{e.g.} $\mathsf{D} - \mathsf{M}$\\
    $\bm{v}$ & The spectral coordinate axis $\bm{\lambda}$ expressed as a velocity difference\\
    \hline
    \multicolumn{2}{c}{Line properties}\\
    \hline
    $\lambda_{\mathrm{c},j}$ & Line center position of the $j^{th}$ spectral line\\
    $a_j$ & Gaussian line profile amplitude of the $j^{th}$ spectral line \\
    $\sigma_j$ & Gaussian line profile scale of the $j^{th}$ spectral line\\
    $\gamma_j$ & Lorentzian line profile half width of the $j^{th}$ spectral line\\
    $\mathsf{V}_j$ & The Voigt profile of the $j^{th}$ spectral line \\
    $\bar{\bm{F}}$ & The dense $(N_{\rm lines} \times N_{x})$ matrix of all line fluxes stacked vertically \\
    $\hat{\bm{F}}$ & The sparse $(N_{\rm lines} \times N_{\rm sparse})$ matrix of all line fluxes stacked vertically \\
    $\zeta$ & The rotational broadening convolution kernel\\
    $g$ & The instrumental broadening convolution kernel, typically a Gaussian\\
    \hline
    \multicolumn{2}{c}{Scalars}\\
    \hline
    $N_{\rm lines}$ & Number of spectral lines \\
    $N_{x}$ & Number of pixel coordinates in the precomputed spectrum $\bm{\lambda}_x$\\
    $N_{\rm sparse}$ & Number of non-zero pixels computed in the sparse implementation\\
    $\pm \Delta \lambda_{\mathrm{buffer}}$ & Buffer exceeding the red and blue limits of the data spectrum\\
    $P_{\rm rom}$ & The prominence threshold of spectral lines to include in cloning \\
    $v\sin{i}$ & Rotational broadening for stellar inclination $i$ and equatorial velocity $v$\\
    $RV$ & Radial velocity of the star\\
    $R$ & Spectrograph resolving power $\lambda/\delta\lambda$\\
    $\mathcal{L}$ & The loss scalar, usually the sum of the squares of the residuals\\
    \hline
    \multicolumn{2}{c}{Operators}\\
    \hline
    $\resample \big[ \mathsf{F(\bm{\lambda}_x)} \big]$ & The resample operator, takes in a flux spectrum $\mathsf{F}$ evaluated at $\bm{\lambda}_x$ coordinates and returns the mean flux within the pixel boundaries of coordinate $\bm{\lambda}_z$\\
    $*$& The convolution operator\\
    $\odot$& \emph{Hadamard product}, an elementwise product of two same-length vectors\\
    \enddata

\end{deluxetable}

>>>>

\documentclass[modern]{aastex631}
\usepackage{hyperref}
\bibliographystyle{aasjournal}
%\turnoffedit
\usepackage[caption=false]{subfig}
\usepackage{booktabs}
\usepackage{censor}
\usepackage{mathtools}
\usepackage{outlines}
\usepackage{amsmath,bm}
\definecolor{belize}{RGB}{41, 128, 185}
\definecolor{peter}{RGB}{52, 152, 219}
\definecolor{nephritis}{RGB}{39, 174, 96}
\definecolor{asbestos}{RGB}{127, 140, 141}
\hypersetup{linkcolor=nephritis,citecolor=belize,filecolor=asbestos,urlcolor=peter}

@@

In extreme precision radial velocity (EPRV) applications, cross-correlation methods work \citep{2018A_A...620A..47D}, but have many limitations \citep{2022arXiv220110639Z}.  Among the many such limitations, one pernicious noise floor stands out as enigmatic: telluric mitigation.  Many practitioners today simply mask these telluric regions, yet micro-tellurics still inject variance into the spectrum that cannot be easily accounted for with existing methods.  A robust accounting of telluric absorption at the $\mathrm{cm/s}$ level may require joint modeling of the star and the Earth's atmospheric absorption \emph{before} convolution with an instrumental kernel.  This telluric joint modeling capability does not yet exist at a precision that can meet these strenuous demands.

@@

In this paper we show that autodiff-aware semi-empirical models offer an appealing middle ground: informed from self-consistent models, but refined with data.  In Section \ref{methodology} we introduce the \emph{blas\'e} forward-model design and its \texttt{PyTorch}-based \citep{2019arXiv191201703P} implementation.  We show how to supplant the computationally expensive radiative transfer by cloning pre-computed synthetic spectral models.  We show how to combine both stellar and telluric models in a joint modelling approach.  In Section \ref{transferLearn} we describe how to fine-adjust the cloned models to real-world data from high-grasp near-IR spectrographs, using a transfer-learning step.
%Section \ref{secInjRec} exhibits injection-recovery tests on simulated data of known properties, revealing the scalability and fundamental limits of the technique.  
Finally we discuss many conceivable extensions to the framework, and directions for future work in Section \ref{secDiscuss}.

@@

At this stage, we have the choice of whether to work in linear or log scale flux units. Adopting the log of the flux would ensure that the cloned model possesses only positive flux values, a desirable trait of any physical spectral model. We have implemented both modes in \texttt{blas\'e}, allowing users to choose their preference.  We only narrate the linear flux unit description in the main text of this document, for the sake of clarity, and since most practitioners tend to think of flux in terms of linear flux units.  The data-model comparison step will always take place in linear flux units, so the only operational difference is the behavior for deep and saturated lines.  Appendix \ref{appendixLogScale} lists the equations adjusted to log flux units.

@@

% Table
\begin{deluxetable}{cp{10cm}}
    \tabletypesize{\scriptsize}
    \tablecaption{Notation used in this paper\label{table2}}
    \tablehead{
        \colhead{Symbol} & \colhead{Meaning}
    }
    \startdata
    \hline
    \multicolumn{2}{c}{Spectra}\\
    \hline
    $\bm{\lambda}_S$ & Native wavelength coordinates of the precomputed stellar spectrum\\
    $\bm{\lambda}_T$ & Native wavelength coordinates of the telluric spectrum\\
    $\bm{\lambda}_D$ & Native wavelength coordinates of the data spectrum\\
    $\mathsf{S}_{\rm abs}$ & Flux values of the precomputed synthetic stellar spectral model $\bm{\lambda}_S$\\
    $\mathsf{B}$ & Blackbody of temperature $T_{\mathrm{eff}}$ to coarsely normalize $\mathsf{S}_{\rm native}$\\
    $\mathsf{P}$ & Smooth polynomial to refine continuum-normalization\\
    $\mathsf{S}$ & Continuum normalized augmentation of $\mathsf{S}_{\rm abs}$\\
    $\mathsf{T}$ & Transmission values of the precomputed synthetic telluric model \\
    $\mathsf{D}$ & The observed data spectrum flux values\\
    $\bm{\epsilon}$ & The estimated uncertainties in the data spectrum\\
    $\mathsf{S}_{\rm clone}$ & Evaluable and tunable cloned flux model of $\mathsf{S}$\\
    $\mathsf{T}_{\rm clone}$ & Evaluable and tunable cloned transmission model of $\mathsf{T}$\\
    $\mathsf{S}_{\rm ext}$ & An augmentation of $\mathsf{S}_{\rm clone}$ with $v\sin{i}$ convolution and $RV$ translation\\
    $\mathsf{M}_{\rm joint}$ & The joint stellar and telluric model: $\mathsf{S}_{\rm ext} \odot \mathsf{T}_{\rm clone}(\bm{\lambda}_S)$  \\
    $\mathsf{M}$ & Joint model convolved
    with instrumental kernel and resampled to $\bm{\lambda}_D$\\
    $\mathsf{R}$ & The residual spectrum between a pair of inputs, \emph{e.g.} $\mathsf{D} - \mathsf{M}$\\
    $\bm{v}$ & The spectral coordinate axis $\bm{\lambda}$ expressed as a velocity difference\\
    \hline
    \multicolumn{2}{c}{Line properties}\\
    \hline
    $\lambda_{\mathrm{c},j}$ & Line center position of the $j^{th}$ spectral line\\
    $a_j$ & Gaussian line profile amplitude of the $j^{th}$ spectral line \\
    $\sigma_j$ & Gaussian line profile scale of the $j^{th}$ spectral line\\
    $\gamma_j$ & Lorentzian line profile half width of the $j^{th}$ spectral line\\
    $\mathsf{V}_j$ & The Voigt profile of the $j^{th}$ spectral line \\
    $\bar{\bm{F}}$ & The dense $(N_{\rm lines} \times N_{x})$ matrix of all line fluxes stacked vertically \\
    $\hat{\bm{F}}$ & The sparse $(N_{\rm lines} \times N_{\rm sparse})$ matrix of all line fluxes stacked vertically \\
    $\zeta$ & The rotational broadening convolution kernel\\
    $g$ & The instrumental broadening convolution kernel, typically a Gaussian\\
    \hline
    \multicolumn{2}{c}{Scalars}\\
    \hline
    $N_{\rm lines}$ & Number of spectral lines \\
    $N_{x}$ & Number of pixel coordinates in the precomputed spectrum $\bm{\lambda}_x$\\
    $N_{\rm sparse}$ & Number of non-zero pixels computed in the sparse implementation\\
    $\pm \Delta \lambda_{\mathrm{buffer}}$ & Buffer exceeding the red and blue limits of the data spectrum\\
    $P_{\rm rom}$ & The prominence threshold of spectral lines to include in cloning \\
    $v\sin{i}$ & Rotational broadening for stellar inclination $i$ and equatorial velocity $v$\\
    $RV$ & Radial velocity of the star\\
    $R$ & Spectrograph resolving power $\lambda/\delta\lambda$\\
    $\mathcal{L}$ & The loss scalar, usually the sum of the squares of the residuals\\
    \hline
    \multicolumn{2}{c}{Operators}\\
    \hline
    $\resample \big[ \mathsf{F(\bm{\lambda}_x)} \big]$ & The resample operator, takes in a flux spectrum $\mathsf{F}$ evaluated at $\bm{\lambda}_x$ coordinates and returns the mean flux within the pixel boundaries of coordinate $\bm{\lambda}_z$\\
    $*$& The convolution operator\\
    $\odot$& \emph{Hadamard product}, an elementwise product of two same-length vectors\\
    \enddata
\end{deluxetable}
